A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-01-17 sections/algorithm.tex
Fix compile error
sections; add algo. rules; few text on algo

$$\begin{aligned}
\text{Type variables}\qquad&a, b\\
\text{Existential variables}\qquad&\a, \b\\[3mm]
\text{Types}\qquad&A', B', C' &::=&\quad 1 \mid a \mid \forall x. A' \mid A'\to B'\\
\text{Mono-types}\qquad&\tau &::=&\quad 1 \mid a \mid \tau_1\to \tau_2\\
\text{Context}\qquad&\P &::=&\quad \cdot \mid \P, a \mid \P, x:A' \color{red} \\[3mm]
% \text{Declarative Judgements}\qquad&\jo &::=&\quad \cdot \mid \P\vdash A' \le B' \jc \jo\\
% &&&\quad \mid \P\vdash e\Leftarrow A' \jc \jo \mid \P\vdash e\To A' \jc \jo \mid \P\vdash A' \bullet e\TTo C' \jc \jo\\
% \noalign{\jimmy{\text{What do the markers do: $\G \mark,a \mark,b\vdash \j_1;\j_2;\j_3$ means $\G,a,b\vdash \j_1 \land \G,a\vdash \j_1,\j_2 \land \G\vdash \j_3$}}}\\[3mm]
\text{Algorithmic types}\qquad&A, B, C &::=&\quad 1 \mid a \mid \a \mid \forall a. A \mid A\to B\\
\text{Algorithmic judgement chain}\qquad&\j &::=&\quad A \le B \mid e\Lto A \mid e\To_{a} \j \mid \appInfAlg{A}{e}{a}{\j}\\[3mm]
\text{Algorithmic worklist}\qquad&\G &::=&\quad \cdot \mid \G, a \mid \G, \a \mid \G, x: A \mid \G \mark \j\\[3mm]
\text{Terms}\qquad&e&::=&\quad x \mid () \mid \lam{x}{e} \mid e_1~e_2 \mid (e:A)
\end{aligned}$$

>>>>

\subsection{Syntax and Well-Formedness}

$$\begin{aligned}
\text{Type variables}\qquad&a, b\\
\text{Existential variables}\qquad&\al, \bt\\[3mm]
%\text{Types}\qquad&A', B', C' &::=&\quad 1 \mid a \mid \forall x. A' \mid A'\to B'\\
\text{Mono-types}\qquad&\tau &::=&\quad 1 \mid a \mid \tau_1\to \tau_2\\
%\text{Context}\qquad&\Psi &::=&\quad \cdot \mid \Psi, a \mid \Psi, x:A' \color{red} \\[3mm]
% \text{Declarative Judgements}\qquad&\jo &::=&\quad \cdot \mid \P\vdash A' \le B' \jc \jo\\
% &&&\quad \mid \P\vdash e\Leftarrow A' \jc \jo \mid \P\vdash e\To A' \jc \jo \mid \P\vdash A' \bullet e\TTo C' \jc \jo\\
% \noalign{\jimmy{\text{What do the markers do: $\G \vdash,a \vdash,b\vdash \j_1;\j_2;\j_3$ means $\G,a,b\vdash \j_1 \land \G,a\vdash \j_1,\j_2 \land \G\vdash \j_3$}}}\\[3mm]
\text{Algorithmic types}\qquad&A, B, C &::=&\quad 1 \mid a \mid \al \mid \forall a. A \mid A\to B\\
\text{Algorithmic judgement chain}\qquad&\jg &::=&\quad A \le B \mid e\Lto A \mid e\To_{a} \jg \mid \appInfAlg{A}{e}\\[3mm]
\text{Algorithmic worklist}\qquad&\Gm &::=&\quad \cdot \mid \Gm, a \mid \Gm, \al \mid \Gm, x: A \mid \Gm \vdash \jg\\[3mm]
\text{Terms}\qquad&e&::=&\quad x \mid () \mid \lam{e} \mid e_1~e_2 \mid (e:A)
\end{aligned}$$

\paragraph{Existential Variables} ...

\paragraph{Judgment and judgment chain} ...
\paragraph{Worklist Judgments} Our algorithmic judgment has a non-standard form.
We combine the traditional contexts and judgment(s) into a single worklist.
The worklist collects variable bindings and judgments in order, which represents their scopings:
each variable is visible after it is declared.
For example, $a,x:a,x \Lto a$ is a valid worklist, but $\textcolor{red}{x} \Lto a, x:\textcolor{red}{a},a$ is not.

\paragraph{Hole Notation}...

\begin{figure}[t]
\framebox{$\Gm\vdash\jg$} Well-formed algorithmic judgment.
\begin{gather*}
\inferrule*[right=$\mathtt{wf{\le}}$]
{\Gm\vdash A \\ \Gm\vdash B}
{\Gm\vdash A\le B}
\qquad
\inferrule*[right=$\mathtt{wf{\Lto}}$]
{\Gm\vdash e \\ \Gm\vdash A}
{\Gm\vdash e \Lto A}
\\
\inferrule*[right=$\mathtt{wf{\To}}$]
{\Gm\vdash e \\ \Gm, a\vdash \jg}
{\Gm\vdash e \To_a \jg}
\qquad
\inferrule*[right=$\mathtt{wf{\TTo}}$]
{\Gm\vdash A \\ \Gm, a\vdash \jg \\ \Gm\vdash e}
{\Gm\vdash \appInfAlg{A}{e}}
\end{gather*}

\framebox{$\text{wf }\Gm$} Well-formed algorithmic worklist.
\begin{gather*}
\inferrule*[right=$\mathtt{wf\cdot}$]
{~}{\text{wf }\cdot}
\qquad
\inferrule*[right=$\mathtt{wf_a}$]
{\text{wf }\Gm}
{\text{wf }\Gm, a}
\qquad
\inferrule*[right=$\mathtt{wf_{\al}}$]
{\text{wf }\Gm}
{\text{wf }\Gm, \al}
\qquad
\inferrule*[right=$\mathtt{wf_{of}}$]
{\text{wf }\Gm \\ \Gm\vdash A}
{\text{wf }\Gm, x:A}
\qquad
\inferrule*[right=$\mathtt{wf_{\jg}}$]
{\text{wf }\Gm \\ \Gm\vdash\jg}
{\text{wf }\Gm \vdash\jg}
\end{gather*}
\caption{Algorithmic Well-Formedness}\label{fig:alg}
\end{figure}

\subsection{Algorithmic System}

% TODO rules too long
\begin{figure}[t]
\framebox{$\Gm\rto \Gm'$} $\Gm$ reduces to $\Gm'$.
%\jimmy{Whenever there is a change on judgment count, marker should be properly added/removed. Decidability: all valid judgment chain should be reduced to nil.}
\begin{gather*}
\begin{aligned}
\Gm, a &\rto \Gm \qquad\qquad
\Gm, \al \rto \Gm \qquad\qquad
\Gm, x:A \rto \Gm
\\[2mm]
\Gm \vdash 1\le 1 &\rto \Gm\\
\Gm \vdash a\le a &\rto \Gm\\
\Gm \vdash \al\le \al &\rto \Gm
\\[2mm]
\Gm \vdash A_1\to A_2 \le B_1\to B_2 &\rto \Gm \vdash A_2 \le B_2 \vdash B_1\le A_1\\
\Gm \vdash \forall a. A\le B &\rto \Gm,\al \vdash [\al/a]A\le B \quad\text{\jimmy{No more existential variable "leaks"!}}\\
\Gm \vdash A\le \forall b. B &\rto \Gm,b \vdash A\le B \quad\text{\jimmy{And no more type variable "leaks"!}}
\\[2mm]
%\\
%\text{Let } \color{red}\Gm[\al \toto B // G_M] &:= \Gm_L, G_M, [B/\al]\Gm_R, \text{ when } \Gm[\al] = \Gm_L,\al,\Gm_R \ (G_M\text{ defaults to }\cdot)\\
\Gm[\al] \vdash \al \le A\to B &\rto [\al[1]\to\al[2]/\al] (\Gm[\al[1], \al[2]] \vdash \al[1]\to \al[2] \le A \to B)\\
 &\qquad \text{when }\al\notin FV(A)\cup FV(B)\\
\Gm[\al] \vdash A\to B \le \al &\rto [\al[1]\to \al[2]/\al] (\Gm[\al[1], \al[2]] \vdash A \to B \le \al[1]\to \al[2])\\
 &\qquad \text{when }\al\notin FV(A)\cup FV(B)
 \\[2mm]
\Gm[\al][\bt] \vdash \al \le \bt &\rto [\al/\bt](\Gm[\al][])\\
\Gm[\al][\bt] \vdash \bt \le \al &\rto [\al/\bt](\Gm[\al][])\\
\Gm[a][\bt] \vdash a \le \bt &\rto [a/\bt](\Gm[a][])\\
\Gm[a][\bt] \vdash \bt \le a &\rto [a/\bt](\Gm[a][])\\
\Gm[\bt] \vdash 1 \le \bt &\rto [1/\bt](\Gm[])\\
\Gm[\bt] \vdash \bt \le 1 &\rto [1/\bt](\Gm[])
\\[2mm]
\Gm \vdash e \Lto B &\rto \Gm \vdash e\To_a a\le B\\
\Gm \vdash () \Lto 1 &\rto \Gm\\
\Gm \vdash e\Lto \forall a. A &\rto \Gm,a \vdash e\Lto A\\
\Gm \vdash \lambda x. e \Lto A\to B &\rto \Gm, x:A  \vdash e \Lto B\\
\Gm[\al] \vdash \lambda x. e \Lto \al &\rto [\al[1]\to \al[2] / \al](\Gm[\al[1],\al[2]], x:\al[1] \vdash e \Lto \al[2]) \quad\text{\jimmy{Additional case}}
\\[2mm]
\Gm \vdash x\To_a \jg &\rto \Gm \vdash [A/a] \jg \qquad (x:A)\in \Gm\\
\Gm \vdash (e:A)\To_a \jg &\rto \Gm \vdash [A/a]\jg \vdash e \Lto A\\
\Gm \vdash ()\To_a \jg &\rto \Gm \vdash [1/a]\jg \qquad\jimmy{\downarrow\text{Precise control on exvars: to the $2^{nd}$ chain}}\\
\Gm \vdash \lambda x. e \To_a \jg &\rto
	\Gm,\al,\bt \vdash [\al\to\bt/a]\jg, x:\al \vdash e\Lto \bt \quad\jimmy{\text{differs from Dunfield's}}\\
\Gm \vdash e_1\ e_2 \To_a \jg &\rto \Gm \vdash e_1\To_{a_1} (\appInfAlg{a_1}{e_2})
\\[2mm]
\Gm \vdash \appInfAlg{\forall a. A}{e} &\rto \Gm,\al \vdash \appInfAlg{[\al/a]A}{e}\\
\Gm \vdash \appInfAlg{A\to C}{e} &\rto \Gm \vdash [C/a]\jg \vdash e \Lto A\\
\Gm[\al] \vdash \appInfAlg{\al}{e} &\rto
	[\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \vdash [\al[2]/a]\jg \vdash e\Lto \al[1])\\
 &\color{magenta} \makebox[0pt]{\qquad or}\phantom{{}\rto{}}
 	[\al[1]\to\al[2]/\al](\Gm[\al[1], \al[2]] \vdash \appInfAlg{\al[1]\to\al[2]}{e}) \quad\text{(1 step equiv.)}
\end{aligned}
\end{gather*}
\caption{Algorithmic Typing}\label{fig:alg}
\end{figure}
